\Title

### 2013-05-17 synmut.tex
Small changes in text

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Template for PLoS
% Version 1.0 January 2009
\documentclass[10pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{color} 
% Use doublespacing - comment out for single spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm

% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

% Use the PLoS provided bibtex style
\bibliographystyle{plos2009}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

\pagestyle{myheadings}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{eucal,eurosym,textcomp}
\usepackage[caption=false]{subfig}
\usepackage{pslatex}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\pfix}{P_{\mathrm{fix}}}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\shankaregion}{C2-V5}
\newcommand{\PCApat}{1}
\newcommand{\timedependence}{3}
\newcommand{\withinepi}{4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV-1 \env~evolution}
\newcommand{\Affiliation}{Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\begin{flushleft}
{\Large
\textbf{\Title}
}
\\
Fabio~Zanini$^{1}$ and
Richard~A.~Neher$^{1,\ast}$
\\
\bf{1} \Affiliation
\\
$\ast$ E-mail: richard.neher@tuebingen.mpg.de
\end{flushleft}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Please keep the abstract between 250 and 300 words
\section*{Abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Intrapatient HIV-1 evolution is dominated by selection on the protein level in the
arms race with the adaptive immune system. When cytotoxic CD8${}^+$ T-cells or 
neutralizing antibodies
target a new epitope, the virus often escapes via nonsynonymous mutations that
impair recognition. Synonymous mutations do not affect this interplay and are
often assumed to be neutral.
We analyze longitudinal intrapatient data from the \shankaregion{} part of the
envelope gene (\env{}) and observe that synonymous derived alleles rarely
fix even though they often reach high frequencies in the viral population.
We find that synonymous mutations that disrupt base pairs in RNA stems flanking
the variable loops of gp120 are more likely to be lost than other synonymous
changes, hinting at a direct fitness effect of these stem-loop structures in the
HIV-1 RNA.
Computational modeling indicates that these synonymous mutations have a
(Malthusian) selection coefficient of the order of $-0.002$ and that they are
brought up to high frequency by hitchhiking on neighboring beneficial
nonsynonymous alleles.
The patterns of fixation of nonsynonymous mutations estimated from the
longitudinal data and comparisons with computer models
suggest that escape mutations in \shankaregion{} are only transiently
beneficial, either because the immune system is catching up or because of
competition between equivalent escapes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.   
\section*{Author Summary}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\comment{to be written\dots}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
HIV-1 evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
(nAbs)~\cite{rambaut_causes_2004} and facilitated by the high mutation rate
~\cite{mansky_lower_1995,abram_nature_2010}. When the host develops a CTL or
nAb response against a particular HIV-1 epitope, mutations in the viral genome that
reduce or prevent recognition of the epitope frequently emerge. Escape mutations
in epitopes targeted by CTLs typically evolve during early infection and spread
rapidly through the population~\cite{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving parts of the HIV-1 genome are the variable
loops V1-V5 in the envelope protein gp120, which change to avoid recognition by
nAbs. Escape mutations in \env, the gene encoding gp120, spread through the
population within a few months.
Consistent with this time scale, it is found that serum from a particular time
typically neutralizes virus extracted more than 3-6 months earlier but not contemporary
virus \cite{richman_rapid_2003}.

Escape mutations are selected because they change the amino acid sequence
of viral proteins in a way that reduces antibody binding or epitope presentation. 
Conversely, synonymous mutations are commonly used as
approximately neutral markers in studies of viral evolution. Neutral markers are
very useful since their dynamics can be compared to that of putatively
functional sites to detect purifying or directional selection
\cite{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
maintaining protein function and avoiding the adaptive immune recognition,
however, the HIV-1 genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate
at the RNA level and are sensitive to synonymous changes since these processes
often depend on RNA folding. For example, the HIV-1 \rev{} response element (RRE)
in \env{} enhances nuclear export of full length or partially spliced viral
transcripts via a complex stem-loop RNA structure~\cite{fernandes_hiv-1_2012}.
Another well studied case is the interaction between viral reverse
transcriptase, viral ssRNA, and the host tRNA$^\text{Lys3}$: the latter is
required for priming reverse transcription (RT) and is bound by a pseudoknotted
RNA structure in the viral 5' untranslated region~\cite{barat_interaction_1991,
paillart_vitro_2002}.

Even in the absence of important RNA structures, synonymous codons do not evolve
completely neutrally. Some codons are favored over others in many species
\cite{plotkin_synonymous_2011}. Recent studies have shown that genetically
engineered HIV-1 strains with altered codon usage can in some cases produce more
viral protein, but in general replicate less efficiently
\cite{ngumbela_quantitative_2008, li_codon-usage-based_2012,
keating_rich_2009}. Codon deoptimization has been suggested as an attenuation
strategy for polio and influenza~\cite{mueller_live_2010,coleman_virus_2008}.
Purifying selection beyond the protein sequence is therefore expected
\cite{forsdyke_reciprocal_1995,snoeck_mapping_2011}, and it has been shown that
rates of evolution at synonymous sites vary along the HIV-1 genome
\cite{mayrose_towards_2007}. Positive selection through the host adaptive
immune system, however, is restricted to changes in the amino acid sequence.

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that a substantial fraction of these mutations is deleterious.  We
argue that synonymous mutations reach high frequencies via genetic hitchhiking
due to limited recombination in HIV-1 populations~\cite{neher_recombination_2010,
batorsky_estimate_2011}. We then compare our observations to computational
models of HIV-1 evolution and derive estimates for the effect of synonymous mutations
 on fitness.  Extending the analysis of fixation probabilities to the
nonsynonymous mutations, we show that time-dependent selection or strong
competition of escape mutations inside the same epitope are necessary to explain
the observed patterns of fixation and loss.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency. A neutral mutation
segregating at frequency $\nu$ has a probability $\pfix(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e. with
probability $1-\nu$, it goes extinct. As illustrated in the inset of
\FIG{aft}A, this is a  consequence of the fact that (i) exactly one
individual in the current population will be the common ancestor of the entire
future population at a particular site and (ii) this ancestor has a probability
$\nu$ of carrying the mutation (assuming the neutral mutation is not
preferentially associated with genomes of high or low fitness). Deleterious or
beneficial mutations fix less or more often than neutral ones, respectively.
\FIG{aft} shows the time course of the frequencies of all synonymous and
nonsynonymous mutations observed in \env, \shankaregion, in patient
p10~\cite{shankarappa_consistent_1999}. Despite many synonymous
mutations reaching high frequency, few fix (panel~\ref{fig:aft}A); in
constrast, many nonsynonymous mutations fix (panel~\ref{fig:aft}B).
Strictly speaking, no mutation in the HIV-1 population ever fixes because the 
mutation rate and the population size are large. Therefore, we define ``fixation'' or ``loss''
by not observing the mutation in the sample.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Synonymous polymorphisms in \env, \shankaregion, are mostly deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We study the dynamics and fate of synonymous mutations more quantitatively by
analyzing data from seven patients from 
\cite{shankarappa_consistent_1999} and \cite{liu_selection_2006} as well as three patients from
\cite{bunnik_autologous_2008} (patients whose viral population was structured
were excluded from the analysis; see methods and \figurename~S\PCApat).  The
former data set is restricted to the \shankaregion{} region of \env, while
the data from \cite{bunnik_autologous_2008} cover the majority of \env.
Considering all mutations in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
$t$, we calculate the fraction that are still observed at later times $t+\Delta t$. Plotting this fraction against the time interval $\Delta t$, we see that
most synonymous mutations segregate for roughly one year and are lost much more
frequently than expected (panel \ref{fig:fixp}A). The long-time probability of
fixation, $\pfix$, is shown as a function of the
initial frequency $\nu_0$ in panel~\ref{fig:fixp}B (red line). 
We find that $\pfix$ of synonymous mutations is far below
the neutral expectation in  \shankaregion.  Outside of \shankaregion, using data from
\cite{bunnik_autologous_2008} only, we find no such reduction in $\pfix$.
Restricted to the \shankaregion{} region, the sequence samples from
\cite{bunnik_autologous_2008} are fully compatible with data from
\cite{shankarappa_consistent_1999}. The nonsynonymous mutations seem to follow
more or less the neutral expectation (blue line) -- a point to which we will
come back below.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Synonymous mutations in \shankaregion{} tend to disrupt conserved RNA stems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One possible explanation for lack of fixation of synonymous mutations in
\shankaregion{} is secondary structures in the viral RNA, the disruption of which
is deleterious to the virus \cite{forsdyke_reciprocal_1995,
snoeck_mapping_2011, sanjuan_interplay_2011}.

The propensity of nucleotides in the HIV-1 genome to form base pairs has been
measured using the SHAPE assay, a biochemical reaction preferentially altering
unpaired bases (the HIV-1 genome is a single stranded RNA)
\cite{watts_architecture_2009}. The SHAPE assay has shown that the variable
regions V1-V5 tend to be unpaired, while the conserved regions between those
variable regions form stems.  We aligned the within-patient sequence samples 
to the reference NL4-3 strain used in \cite{watts_architecture_2009} and 
thereby assigned SHAPE reactivities to most positions in the alignment. 
We then calculated the distributions of SHAPE reactivities for synonymous 
polymorphisms that fixed or were subsequently lost (only polymorphisms with 
frequencies above 15\%).
As shown in \FIG{SHAPE}A, the reactivities of fixed alleles (red
histogram) are systematically larger than those of alleles that are lost (blue)
(Kolmogorov-Smirnov test on the cumulative distribution, $p\approx 0.002$). In
other words, alleles that are likely to break RNA helices are also more likely
to revert and finally be lost from the population. The average over all
mutations that are not observed (green) lies between  those that fix and
those that get lost. Note that this analysis will be sensitive only at positions
where the base pairing pattern of NL4-3 agrees with that of each patient's
initial consensus sequence (it is thus statistically conservative).

To test the hypothesis that mutations in \shankaregion{} are lost because they
break stems in the conserved stretches between the variable loops, we consider
mutations in variable loops and conserved parts separately. The greatest
depression in fixation probability is observed in the conserved stems, while the
variable loops show little deviation from the neutral signature, see
\FIG{SHAPE}B. This is consistent with important stem structures in conserved
regions between loops.

In addition to RNA secondary structure, we have considered other possible
explanations for a fitness cost of some synonymous mutations, in particular codon
usage bias (CUB). HIV-1 is known to prefer A-rich codons over highly expressed
human codons~\cite{jenkins_extent_2003,kuyl_biased_2012}. We do not find,
however, any evidence for a contribution of average CUB to the ultimate fate of
synonymous alleles; consistently, HIV-1 does not seem to adapt its codon usage to
its human host cells at the macroevolutionary level \cite{kuyl_biased_2012}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Deleterious mutations are brought to high frequency by hitchhiking}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency and that the fixation probability is reduced only in parts of the
genome (in \shankaregion{} but not in the rest of \env{}; compare the red
triangle line versus the green square line in \FIG{fixp}B).
The region \shankaregion{} undergoes frequent adaptive changes to evade
recognition by neutralizing antibodies \cite{williamson_adaptation_2003,
richman_rapid_2003}. Due to the limited amount of recombination in HIV-1
\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious mutations
that are linked to adaptive variants can reach high frequency. This process is
known as hitchhiking \cite{smith_hitch-hiking_1974} or genetic draft
\cite{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is  apparent in
\FIG{aft}, which shows that many mutations change rapidly in frequency as a
flock.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp, twocolumn]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{natbib}
\usepackage{pslatex}
%\usepackage[colorlinks,linkcolor=red,citecolor=blue]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\pfix}{P_{\mathrm{fix}}}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\shankaregion}{C2-V5}
\newcommand{\PCApat}{1}
\newcommand{\timedependence}{3}
\newcommand{\withinepi}{4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV-1 \env~evolution}
\newcommand{\Affiliation}{Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
%\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\title{\Title}
\author{\Author}
\affiliation{\Affiliation}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}
\noindent
Intrapatient HIV-1 evolution is dominated by selection on the protein level in the
arms race with the adaptive immune system. When cytotoxic CD8${}^+$ T-cells or 
neutralizing antibodies
target a new epitope, the virus often escapes via nonsynonymous mutations that
impair recognition. Synonymous mutations do not affect this interplay and are
often assumed to be neutral.
We analyze longitudinal intrapatient data from the \shankaregion{} part of the
envelope gene (\env{}) and observe that synonymous derived alleles rarely
fix even though they often reach high frequencies in the viral population.
We find that synonymous mutations that disrupt base pairs in RNA stems flanking
the variable loops of gp120 are more likely to be lost than other synonymous
changes, hinting at a direct fitness effect of these stem-loop structures in the
HIV-1 RNA.
Computational modeling indicates that these synonymous mutations have a
(Malthusian) selection coefficient of the order of $-0.002$ and that they are
brought up to high frequency by hitchhiking on neighboring beneficial
nonsynonymous alleles.
The patterns of fixation of nonsynonymous mutations estimated from the
longitudinal data and comparisons with computer models
suggest that escape mutations in \shankaregion{} are only transiently
beneficial, either because the immune system is catching up or because of
competition between equivalent escapes.

\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
HIV-1 evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
(nAbs)~\citep{rambaut_causes_2004} and facilitated by the high mutation rate
~\citep{mansky_lower_1995,abram_nature_2010}. When the host develops a CTL or
nAb response against a particular HIV-1 epitope, mutations in the viral genome that
reduce or prevent recognition of the epitope frequently emerge. Escape mutations
in epitopes targeted by CTLs typically evolve during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving parts of the HIV-1 genome are the variable
loops V1-V5 in the envelope protein gp120, which change to avoid recognition by
nAbs. Escape mutations in \env, the gene encoding gp120, spread through the
population within a few months.
Consistent with this time scale, it is found that serum from a particular time
typically neutralizes virus extracted more than 3-6 months earlier but not contemporary
virus \citep{richman_rapid_2003}.

Escape mutations are selected because they change the amino acid sequence
of viral proteins in a way that reduces antibody binding or epitope presentation. 
Conversely, synonymous mutations are commonly used as
approximately neutral markers in studies of viral evolution. Neutral markers are
very useful since their dynamics can be compared to that of putatively
functional sites to detect purifying or directional selection
\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
maintaining protein function and avoiding the adaptive immune recognition,
however, the HIV-1 genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate
at the RNA level and are sensitive to synonymous changes since these processes
often depend on RNA folding. For example, the HIV-1 \rev{} response element (RRE)
in \env{} enhances nuclear export of full length or partially spliced viral
transcripts via a complex stem-loop RNA structure~\citep{fernandes_hiv-1_2012}.
Another well studied case is the interaction between viral reverse
transcriptase, viral ssRNA, and the host tRNA$^\text{Lys3}$: the latter is
required for priming reverse transcription (RT) and is bound by a pseudoknotted
RNA structure in the viral 5' untranslated region~\citep{barat_interaction_1991,
paillart_vitro_2002}.

Even in the absence of important RNA structures, synonymous codons do not evolve
completely neutrally. Some codons are favored over others in many species
\citep{plotkin_synonymous_2011}. Recent studies have shown that genetically
engineered HIV-1 strains with altered codon usage can in some cases produce more
viral protein, but in general replicate less efficiently
\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012,
keating_rich_2009}. Codon deoptimization has been suggested as an attenuation
strategy for polio and influenza~\citep{mueller_live_2010,coleman_virus_2008}.
Purifying selection beyond the protein sequence is therefore expected
\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}, and it has been shown that
rates of evolution at synonymous sites vary along the HIV-1 genome
\citep{mayrose_towards_2007}. Positive selection through the host adaptive
immune system, however, is restricted to changes in the amino acid sequence.

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that a substantial fraction of these mutations is deleterious.  We
argue that synonymous mutations reach high frequencies via genetic hitchhiking
due to limited recombination in HIV-1 populations~\citep{neher_recombination_2010,
batorsky_estimate_2011}. We then compare our observations to computational
models of HIV-1 evolution and derive estimates for the effect of synonymous mutations
 on fitness.  Extending the analysis of fixation probabilities to the
nonsynonymous mutations, we show that time-dependent selection or strong
competition of escape mutations inside the same epitope are necessary to explain
the observed patterns of fixation and loss.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency. A neutral mutation
segregating at frequency $\nu$ has a probability $\pfix(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e. with
probability $1-\nu$, it goes extinct. As illustrated in the inset of
\FIG{aftsyn}, this is a  consequence of the fact that (i) exactly one
individual in the current population will be the common ancestor of the entire
future population at a particular site and (ii) this ancestor has a probability
$\nu$ of carrying the mutation (assuming the neutral mutation is not
preferentially associated with genomes of high or low fitness). Deleterious or
beneficial mutations fix less or more often than neutral ones, respectively.
\FIG{aft} shows the time course of the frequencies of all synonymous and
nonsynonymous mutations observed in \env, \shankaregion, in patient
p10~\citep{shankarappa_consistent_1999}. Despite many synonymous
mutations reaching high frequency, few fix (panel~\ref{fig:aftsyn}); in
constrast, many nonsynonymous mutations fix (panel~\ref{fig:aftnonsyn}).
Strictly speaking, no mutation in the HIV-1 population ever fixes because the 
mutation rate and the population size are large. Therefore, we define ``fixation'' or ``loss''
by not observing the mutation in the sample.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous polymorphisms in \env, \shankaregion, are mostly deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We study the dynamics and fate of synonymous mutations more quantitatively by
analyzing data from seven patients from
\citet{shankarappa_consistent_1999} and \citep{liu_selection_2006} as well as three patients from
\citet{bunnik_autologous_2008} (patients whose viral population was structured
were excluded from the analysis; see methods and \figurename~S\PCApat).  The
former data set is restricted to the \shankaregion{} region of \env, while
the data from \citet{bunnik_autologous_2008} cover the majority of \env.
Considering all mutations in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
$t$, we calculate the fraction that are still observed at later times $t+\Delta t$. Plotting this fraction against the time interval $\Delta t$, we see that
most synonymous mutations segregate for roughly one year and are lost much more
frequently than expected (panel \ref{fig:fixp1}). The long-time probability of
fixation, $\pfix$, is shown as a function of the
initial frequency $\nu_0$ in panel~\ref{fig:fixp2} (red line). 
We find that $\pfix$ of synonymous mutations is far below
the neutral expectation in  \shankaregion.  Outside of \shankaregion, using data from
\citet{bunnik_autologous_2008} only, we find no such reduction in $\pfix$.
Restricted to the \shankaregion{} region, the sequence samples from
\citet{bunnik_autologous_2008} are fully compatible with data from
\citet{shankarappa_consistent_1999}. The nonsynonymous mutations seem to follow
more or less the neutral expectation (blue line) -- a point to which we will
come back below.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]
{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
\label{fig:aftsyn}}\\
 \subfloat{\includegraphics[width=\linewidth]
{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of frequencies
of synonymous (A) and nonsynonymous (B) derived alleles in \env, 
\shankaregion, from patient 10~\cite{shankarappa_consistent_1999}.
While many nonsynonymous mutations  fix, few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the \shankaregion{} region
(blue to red). Inset: the fixation probability $\pfix$ of a neutral
mutation is simply the likelihood that the future common ancestor at this 
position is currently carrying it, i.e. the mutation frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_times.pdf}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities.pdf}
\label{fig:fixp2}}
\caption{Fixation and loss of synonymous mutations.
Panel A) shows how quickly synonymous mutations are purged from the populations. 
Specifically, the figure shows the fraction of mutations that are still observed
after $\Delta t$ days, conditional on being observed in one of the three frequency 
intervals (different colors). 
In each frequency interval, the fraction of synonymous
mutations that ultimately survive is the fixation probability $\pfix$ conditional on the
initial frequency. The neutral expectation for $\pfix=\nu_0$ is indicated by 
dashed horizontal lines.
Panel B) shows the fixation probability of derived synonymous
alleles as a function of $\nu_0$. Polymorphisms within \shankaregion{} fix less
often than expected for neutral mutations (indicated by the diagonal line).
This suppression is not observed in other parts of \env~or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
refs.~\cite{shankarappa_consistent_1999,liu_selection_2006, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous mutations in \shankaregion{} tend to disrupt conserved RNA stems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One possible explanation for lack of fixation of synonymous mutations in
\shankaregion{} is secondary structures in the viral RNA, the disruption of which
is deleterious to the virus \citep{forsdyke_reciprocal_1995,
snoeck_mapping_2011, sanjuan_interplay_2011}.

The propensity of nucleotides in the HIV-1 genome to form base pairs has been
measured using the SHAPE assay, a biochemical reaction preferentially altering
unpaired bases (the HIV-1 genome is a single stranded RNA)
\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
regions V1-V5 tend to be unpaired, while the conserved regions between those
variable regions form stems.  We aligned the within-patient sequence samples 
to the reference NL4-3 strain used by \citet{watts_architecture_2009} and 
thereby assigned SHAPE reactivities to most positions in the alignment. 
We then calculated the distributions of SHAPE reactivities for synonymous 
polymorphisms that fixed or were subsequently lost (only polymorphisms with 
frequencies above 15\%).
As shown in \FIG{SHAPEA}, the reactivities of fixed alleles (red
histogram) are systematically larger than those of alleles that are lost (blue)
(Kolmogorov-Smirnov test on the cumulative distribution, $p\approx 0.002$). In
other words, alleles that are likely to break RNA helices are also more likely
to revert and finally be lost from the population. The average over all
mutations that are not observed (green) lies between  those that fix and
those that get lost. Note that this analysis will be sensitive only at positions
where the base pairing pattern of NL4-3 agrees with that of each patient's
initial consensus sequence (it is thus statistically conservative).

To test the hypothesis that mutations in \shankaregion{} are lost because they
break stems in the conserved stretches between the variable loops, we consider
mutations in variable loops and conserved parts separately. The greatest
depression in fixation probability is observed in the conserved stems, while the
variable loops show little deviation from the neutral signature, see
\FIG{SHAPEB}. This is consistent with important stem structures in conserved
regions between loops.

In addition to RNA secondary structure, we have considered other possible
explanations for a fitness cost of some synonymous mutations, in particular codon
usage bias (CUB). HIV-1 is known to prefer A-rich codons over highly expressed
human codons~\citep{jenkins_extent_2003,kuyl_biased_2012}. We do not find,
however, any evidence for a contribution of average CUB to the ultimate fate of
synonymous alleles; consistently, HIV-1 does not seem to adapt its codon usage to
its human host cells at the macroevolutionary level \citep{kuyl_biased_2012}.
